---
title: "Fine-tuning Large-Language Models for better summarization of Ecological Literature"
author:
  - name: Hammed A. Akande
    email: akandehammedadedamola@gmail.com
    correspondingauthor: true
    affiliation: 1
    
  - name: Logan Hysen
    affiliation: 2
    
  - name: Danial Nayeri
    affiliation: 3

  - name: Chris Collier
    affiliation: 4
affiliation:
  - id: 1
    name: "Laboratory of Community and Quantitative Ecology, Department of Biology, Concordia University, Montreal, Canada"
  - id: 2
    name: "Center for Systems Integration and Sustainability, Department of Fisheries and Wildlife, Michigan State University, East Lansing, MI, USA"
  - id: 3
    name: "Department of Wildlife and Fisheries Sciences, Texas A&M University, College Station, TX, USA"
  - id: 4
    name: "Department of Biological Sciences, California Polytechnic State University, Humboldt, CA, USA"
    
    
address:
  - code: "1"
    organization: "Laboratory of Community and Quantitative Ecology, Department of Biology, Concordia University, Montreal, Canada"
    addressline: "7141 Sherbrooke St W."
    city: "Montreal"
    state: "Quebec"
    postcode: "H4B 1R6"
    country: "Canada"
  - code: "2"
    organization: "Center for Systems Integration and Sustainability, Department of Fisheries and Wildlife, Michigan State University, East Lansing, MI, USA"
  - code: "3"
    organization: "Department of Wildlife and Fisheries Sciences, Texas A&M University, College Station, TX, USA"
  - code: "4"
    organization: "Department of Biological Sciences, California Polytechnic State University, Humboldt, CA, USA"
        

abstract: |
  This is the abstract section
  
keywords: 
  - keyword1
  - keyword2
  - keyword3
  - keyword4
journal: "to be decided"
date: "`r Sys.Date()`"
linenumbers: false
numbersections: true
bibliography: mybibfile.bib
biblio-style: elsarticle-harv # author year style for natbib - use 'elsarticle-num' or 'elsarticle-num-names' for numbered scheme
classoption: preprint, 3p, authoryear # remove authoryear is not using `elsarticle-harv`
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

# Introduction

The body of ecological research is growing rapidly, with the number of articles published each year greater than the last. This rapid growth makes it increasingly difficult for researchers to keep up with advances in the field. Systematic literature reviews are one way to synthesize this body of research, but are traditionally completed manually with human reviewers reading abstracts and full texts for hundreds or thousands of articles. Aside from issues with reviewer fatigue and bias, this process is time consuming and is likely to fall behind as the rate of ecology publications continues to increase. This is known as the synthesis gap, which can limit our ability to make connections between research findings, learn about new advances, and make policy recommendations less potent.

While individual articles are important, overall trends in research are particularly useful for policymakers weighing the impact of different decisions. However, the synthesis gap can limit our understanding of those trends, leading to decisions made with insufficient information and inefficient use of resources. Resources are often scarce for both ecology researchers and the agencies that rely on ecology research. Thus, methods that streamline research synthesis need to be adopted by ecologists.

Recent advances in Natural Language Processing (NLP) approaches show promise for making research synthesis more efficient. [describe some of these advances and how they are used, for example in the medical literature and how accurate/time saving they are]. However, many NLP approaches are still being developed and have yet to be implemented at a large scale by ecologists.

In ecology reviews, there are often two types of data collected. One type is information that could easily be classified with little subjectivity. For example, a researcher may wish to describe the years of data collection for a set of studies. A large language model could do this task with little input from the researcher with a high degree of accuracy (cite Gougherty 2024 here). The other type is subjective data that has a complex interpretation. For example, a researcher might want to know what the important questions posed at the end of a set of articles are. For this task, large language models might struggle to classify the data more than for more objective data types. This suggests that NLP might be useful to ecologists, but slightly different approaches might need to be taken depending on the type of data researchers are interested in collecting from the literature.

These approaches can help ecologists shrink the synthesis gap and make better management or conservation recommendations. In this paper, we present a tool developed to leverage cutting-edge NLP approaches to streamline ecology reviews. First, we present the tool and its structure. Then, we demonstrate the tool's use, accuracy, and time-saving capabilities by applying it to a small case study review.

## Using CSL

If `citation_package` is set to `default` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`. In this case, the `csl` option is used to format the references. Alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>. These can be downloaded and stored locally, or the url can be used as in the example header.

# Methods

# Equations

Here is an equation: $$ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
$$

Here is another: \begin{align}
  a^2+b^2=c^2.
\end{align}

Inline equations: $\sum_{i = 2}^\infty\{\alpha_i^\beta\}$

# Figures and tables

Figure \ref{fig2} is generated using an R chunk.

```{r fig2, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig2}A meaningless scatterplot.", echo = FALSE}
plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in Table \ref{tab1} for example.

```{r tab1, echo = TRUE}
knitr::kable(head(mtcars)[,1:4], 
    caption = "\\label{tab1}Caption centered above table"
)
```

# References {.unnumbered}
